### 2017-10-21 src/3_TimeSeries.tex
cap 1 y 2 revisados

 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). 
The returns distribution tails are larger than
 what is hypothesised by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. 
\end{description}

@@

\textbf{Example of spurious regression} \quad 
An example is shown below. Two random walks series $x_{1,t}$ and $x_{2,t}$ with
a small drift ($\alpha=0.09$) are generated and then regressed one on the
other:
 \begin{equation}
 x_{i,t} = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad
\mathcal{N}(0,1), \quad i=1,2
 \end{equation}

\begin{figure}[!h]
  \centering
  \includegraphics[width=0.8\textwidth]{img/spurious1}
  \caption[Random walks time series]{Two random walks time series $x_{1,t}$ and $x_{2,t}$}
  \label{fig:spurious1}
\end{figure}

@@

A (vector) time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation a stationary process is
obtained~\cite{banerjee1993};
more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

@@

\newpage
\textbf{Cointegration example} \quad

@@

@@

This example shows how cointegration vectors describes the stable relations
between the processes by linear relations that are more stationary than the
original process. Cointegration vectors can be obtained using the Johansen method which gives with a certain probability the number of significant vectors. Figures    
\ref{fig:cointvectors3}, \ref{fig:cointvectors4}  and \ref{fig:cointvectors5}  shows the linear combinations obtained using cointegration vectors given by Johansen method. The method confirm that only two vectors are found and the third one given doesn't result in a I(0) process.

\section{Unit root tests}
Many economic and financial time series exhibit trending behaviour or non-stationary behaviour in the mean or variance. An important econometric task is to determine the type of non-stationary. For trend stationary I(0) time series a time-trend regression is appropriate. For I(1) time series taking first differences transforms it into a stationary one.  For cointegration to determine if time series are I(1) is mandatory to model their long-run relationship. In order to determine if a time series is an a I(1) or I(0) process unit root tests are used. Several tests have been developed but the most widely used is an extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented Dickey-Fuller (ADF) test developed by Said and Dickey in 1984\cite{said1984} which tabulated critical values were obtained by~\cite{mackinnon2010}.

@@

\noindent where $\mathbb{D}_t$ is a vector of deterministic terms. Lag length $p$ has to be determined when applying the test, Akaike Information Criteria (AIC) is commonly used to determine it.
The p-lagged difference terms $\Delta y_{t-j}$ are used to approximate an ARMA process and the value of $p$ is set so that the error $\epsilon_t$ is serially uncorrelated. 
Under the null hypothesis, $y_t$ is I(1) which implies that $\phi=1$. 
The test statistic is the following:

@@

\section {Cointegration Tests}

@@

Besides, volatility provides important information for studying asset returns
because the latter are largely uncorrelated and nonlinearly dependent. However,
it has been found that volatility exhibits significant autocorrelation and
predictable patterns~\cite{poon+granger2003} and  many models have been
proposed to forecast its behaviour.

@@

\begin{description}
\item [$\sigma$:] underlying asset price volatility which measures the standard
deviation of the returns
\item [$\mu$:] underlying asset drift which is a measure of the average rate of
growth of the stock
\item[$E$:] option strike or excersice price
\item[$T$:] option date of expiry
\item[$t$:] current time
\item[$r$:] risk-free interest rate
\end{description}

@@

\noindent where $B$ is a Brownian motion. This stochastic differential equation
has two components: a deterministic term given by $\mu S dt$ and a random term
given by $\sigma S dX$.

A brownian motion $B$ (also called a Wiener process) is a stochastic process
characterized by the three following properties:

\begin{description}
\item[Continuity:] $B(t)$ is a continuos function
\item[Normal increments:]  $B(t)-B(s)$ has a normal distribution with mean $0$
and variance $t-s$.
\item[Independence of increments:] for every choice of nonnegative real numbers
$0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random
variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly
independent.
\end{description}

@@

These are known or can be easily obtained from the market, excepting by
volatility which must be estimated. However, rather than assuming a volatility
a priori and computing option prices from it, the model can be used to estimate
volatility at given prices, time to expiration and strike price. This obtained
volatility is called the implied volatility of an option. Additionally, some
models obtain implied volatility from futures (other derivative from prices).

@@

The random walk model, despite its simplicity is still difficult to outperform for standard econometric forecasting models \cite{lo2011} despite its simplicity. The random walk model is defined as:

@@

Notice that, regarding our notation of section (\ref{sec:cointegration}),
we have here 
$\mathbf{y}_t^0 = \mathbf{y}_t$,
$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
the $i$-th component of the vector time series $\mathbf{y}_t^\nu$
is the scalar time series $y_{i,t}^\nu$, where $\nu=1,\dots,p$ and
$i=1,\dots,l$.
Transposing each equation of the system (\ref{eq:var}) we can write
the VAR($p$) model in block-matrix form as:
\begin{equation}\label{eq:simplevareq}
\mathbf{Y} = \mathbf{A} \mathbf{X} + \mathbf{E} \, , 
\end{equation}
%%
\noindent where:
%%
\begin{alignat*}{2}
\mathbf{Y}
&= \begin{bmatrix}
   \mathbf{y}_{p+1}^\top \\
   \mathbf{y}_{p+2}^\top \\
   \vdots \\
   \mathbf{y}_N^\top
   \end{bmatrix}_{(N-p)\times l}
&\quad
\mathbf{A}
&= \begin{pmat}[{...|}]
   \mathbf{y}_p^\top & \mathbf{y}_{p-1}^\top & \dots 
                    & \mathbf{y}_1^\top & 1 \cr
   \mathbf{y}_{p+1}^\top & \mathbf{y}_p^\top & \dots
                       & \mathbf{y}_2^\top & 1 \cr
   \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top & \mathbf{y}_{N-2}^\top & \dots 
                       & \mathbf{y}_{N-p}^\top & 1 \cr
   \end{pmat}_{(N-p)\times(pl+1)} \\
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\Phi}_1^\top \\
   \boldsymbol{\Phi}_2^\top \\
   \vdots \\
   \boldsymbol{\Phi}_p^\top \\
   \mathbf{c}^\top
   \end{bmatrix}_{(pl+1)\times l}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}_{(N-p)\times l}
\end{alignat*}
Taking into account the error term $\mathbf{E}$, equation \ref{eq:simplevareq}
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation (see section \ref{sec:OLS}).

\section{Vector error correction model}\label{sec:vecm}
The VAR model cannot be used with non-stationary variables. VECM ,
developed by \cite{engle87}, is also a linear model for I(1) variables that are
also cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
differences are stationary and they introduce an error correction term which
adjusts coefficients to bring the variables back to equilibrium.

It is obtained re-writing equation \ref{eq:var} in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of the model VAR shown in equation \ref{eq:var}, are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\cite{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorization of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
nonsingular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

@@

Taking into account the error term $\mathbf{E}$, equation~\ref{eq:simpleveceq}
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation (see section \ref{sec:OLS}).

\subsection{Ordinary Least Squares method} \label{sec:OLS}

@@

\newpage
Conventional regression estimators, including VARs, have good properties when
applied to covariance-stationary time series, but encounter difficulties when
applied to non stationary or integrated processes.

These difficulties were illustrated by Granger and Newbold in 1974
\cite{granger1974} when they introduced the concept of spurious regressions
(see section \ref{sec:spurious}). In 1982, Nelson and Plosser \cite{nelson1982}
showed that unit roots might be present in a wide variety of macroeconomic
series in levels or logarithms. This finding gave rise to the industry of unit
root testing, and the implication that variables should be rendered stationary
by differencing before they are included in an econometric model.  Further
theoretical developments by Granger and Engle in 1987 \cite{engle87} raised the
possibility that two or more integrated, non stationary time series might be
cointegrated, so that some linear combination of these series could be
stationary even though each series were not.

If two series are both integrated (of order one, or I(1)) we could model their
interrelationship by taking first differences of each series and including the
differences in a VAR or a structural model.  However, this approach would be
suboptimal if it was determined that these series are indeed cointegrated. In
that case, the VAR only express the short-run responses of these series to
innovations in each series. This implies that the simple regression in first
differences is misspecified.  If the series are cointegrated, they move
together in the long-run. A VAR in first differences will not capture those
long-run tendencies.  VECM fit to the first differences of the non stationary
variables, but a lagged error-correction term is added to the relationship
representing the long-run relationship.

>>>>

 \begin{figure}[h]
 \centering
 \includegraphics[width=0.9\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). 
The returns distribution tails are larger than what is hypothesised by common
data generation process (generally normal distribution assumption). In the
markets, fat tails are an undesirable feature because of the additional risk
they imply.  However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
In the figure \ref{fig:returndist} is shown the SPY returns distribution based
on daily dates from the period 1st July 1998 to 4th April 2013. The distribution
was compared against the normal distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. 
\end{description}

\vspace{15mm}

@@

\vspace{15mm}
\textbf{Example of spurious regression} \quad 
An example is shown below. Two random walks series $x_{1,t}$ and $x_{2,t}$ with
a small drift ($\alpha=0.09$) are generated and then regressed one on the
other:
 \begin{equation}
 x_{i,t} = \alpha + x_{i,t-1} + \epsilon_i,t  \qquad \text{where} \qquad
\mathcal{N}(0,1), \quad i=1,2
 \end{equation}

Figure \ref{fig:spurious1} shows the time series $x_{1,t}$ and $x_{2,t}$ and
figure \ref{fig:spurious2} shows the regression between them and how they seem
to be related.
\begin{figure}[!h]
  \centering
  \includegraphics[width=0.8\textwidth]{img/spurious1}
  \caption[Random walks time series]{Two random walks time series $x_{1,t}$ and $x_{2,t}$}
  \label{fig:spurious1}
\end{figure}

@@

A vector time series $\mathbf{y}_t$ is said to be {\em integrated of order\/}
$d$, and then we shall write $\mathbf{y}_t\sim I(d)$, whenever after $d$ times
(discrete) differentiation a stationary process is
obtained~\cite{banerjee1993};
more precisely, whenever
$(1-L)^d\,\mathbf{y}_t\sim\text{I(0)}$, where $L$ is the usual lag operator:
$(1-L)\,\mathbf{y}_t = \Delta\mathbf{y}_t = \mathbf{y}_t-\mathbf{y}_{t-1}$ for
all $t$.

@@

\vspace{15mm}

\textbf{Cointegration example} \quad

@@

Figure \ref{fig:cointex} shows the three time series examples.

@@

This example shows how cointegration vectors describes the stable relations
between the processes by linear relations that are more stationary than the
original process. Cointegration vectors can be obtained using the Johansen
method which gives with a certain probability the number of significant vectors.
Figures \ref{fig:cointvectors3}, \ref{fig:cointvectors4}  and
\ref{fig:cointvectors5}  shows the linear combinations obtained using
cointegration vectors given by Johansen method. The method confirm that only two
vectors are found and the third one given doesn't result in a I(0) process.
\vfill

\section{Unit root tests}
Many economic and financial time series exhibit trending behaviour or
non-stationary behaviour in the mean or variance. An important econometric task
is to determine the type of non-stationarity. For trend stationary I(0) time
series a time-trend regression is appropriate. For I(1) time series taking first
differences transforms it into a stationary one.  For cointegration to determine
if time series are I(1) is mandatory to model their long-run relationship. In
order to determine if a time series is an a I(1) or I(0) process unit root tests
are used. Several tests have been developed but the most widely used is an
extension of the Dickey and Fuller test \cite{dickey1979} called the Augmented
Dickey-Fuller (ADF) test developed by Said and Dickey in 1984 \cite{said1984}
which tabulated critical values were obtained by~\cite{mackinnon2010}.

@@

\noindent where $\mathbb{D}_t$ is a vector of deterministic terms. Lag length
$p$ has to be determined when applying the test, Akaike Information Criteria
(AIC) is commonly used to determine it.  The p-lagged difference terms $\Delta
y_{t-j}$ are used to approximate an ARMA process and the value of $p$ is set so
that the error $\epsilon_t$ is serially uncorrelated. Under the null hypothesis,
$y_t$ is I(1) which implies that $\phi=1$. The statistic test is the following:

@@

\vfill

@@

Besides, volatility provides important information for studying asset returns
because they largely uncorrelated and nonlinearly dependent. However,
it has been found that volatility exhibits significant autocorrelation and
predictable patterns~\cite{poon+granger2003} and  many models have been
proposed to forecast its behaviour.

@@

\begin{description}
\item [$\sigma$] underlying asset price volatility which measures the standard
deviation of the returns
\item [$\mu$] underlying asset drift which is a measure of the average rate of
growth of the stock
\item[$E$] option strike or excersice price
\item[$T$] option date of expiry
\item[$t$] current time
\item[$r$] risk-free interest rate
\end{description}

@@

\noindent where $B$ is a Brownian motion. This stochastic differential equation
has two components: a deterministic term given by $\mu S dt$ and a random term
given by $\sigma S dB$.

A brownian motion $B$ (also called a Wiener process) is a stochastic process
characterised by the three following properties:

\begin{description}
\item[Continuity] $B(t)$ is a continuos function
\item[Normal increments]  $B(t)-B(s)$ has a normal distribution with mean $0$
and variance $t-s$.
\item[Independence of increments] for every choice of nonnegative real numbers
$0 \leq s_1 <  t_1 \leq \cdots \leq s_n < t_n < \infty$, the increment random
variables $W_{t_1} - W_{s_1}, \cdots, W_{t_n} - W_{s_n}$ are jointly
independent.
\end{description}

@@

@@

The random walk model, despite its simplicity, is still difficult to outperform
for standard econometric forecasting models \cite{lo2011}. The random walk model
is defined as:

@@

Notice that, regarding our notation of the cointegration section
(\ref{sec:cointegration}), we have here 
$\mathbf{y}_t^0 = \mathbf{y}_t$,
$\mathbf{y}_t^\nu = \mathbf{y}_{t-\nu}$ and
the $i$-th component of the vector time series $\mathbf{y}_t^\nu$
is the scalar time series $y_{i,t}^\nu$, where $\nu=1,\dots,p$ and
$i=1,\dots,l$.
Transposing each equation of the system (\ref{eq:var}) we can write
the VAR($p$) model in block-matrix form as:
\begin{equation}\label{eq:simplevareq}
\mathbf{Y} = \mathbf{A} \mathbf{X} + \mathbf{E} \, , 
\end{equation}
%%
\noindent where:
%%
\begin{alignat*}{2}
\mathbf{Y}
&= \begin{bmatrix}
   \mathbf{y}_{p+1}^\top \\
   \mathbf{y}_{p+2}^\top \\
   \vdots \\
   \mathbf{y}_N^\top
   \end{bmatrix}_{(N-p)\times l}
&\quad
\mathbf{A}
&= \begin{pmat}[{...|}]
   \mathbf{y}_p^\top & \mathbf{y}_{p-1}^\top & \dots 
                    & \mathbf{y}_1^\top & 1 \cr
   \mathbf{y}_{p+1}^\top & \mathbf{y}_p^\top & \dots
                       & \mathbf{y}_2^\top & 1 \cr
   \vdots & \vdots & \ddots & \vdots & \vdots \cr
   \mathbf{y}_{N-1}^\top & \mathbf{y}_{N-2}^\top & \dots 
                       & \mathbf{y}_{N-p}^\top & 1 \cr
   \end{pmat}_{(N-p)\times(pl+1)} \\
\mathbf{X}
&= \begin{bmatrix}
   \boldsymbol{\Phi}_1^\top \\
   \boldsymbol{\Phi}_2^\top \\
   \vdots \\
   \boldsymbol{\Phi}_p^\top \\
   \mathbf{c}^\top
   \end{bmatrix}_{(pl+1)\times l}
&\quad
\mathbf{E}
&= \begin{bmatrix}
   \boldsymbol{\epsilon}_{p+1}^\top \\
   \boldsymbol{\epsilon}_{p+2}^\top \\
   \vdots \\
   \boldsymbol{\epsilon}_N^\top \\
   \end{bmatrix}_{(N-p)\times l}
\end{alignat*}
Taking into account the error term $\mathbf{E}$, equation \ref{eq:simplevareq}
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation (see section \ref{sec:OLS}).

Conventional regression estimators, including VARs, have good properties when
applied to covariance-stationary time series, but encounter difficulties when
applied to non stationary or integrated processes.

These difficulties were illustrated by Granger and Newbold in 1974
\cite{granger1974} when they introduced the concept of spurious regressions (see
section \ref{sec:spurious}). In 1982, Nelson and Plosser \cite{nelson1982}
showed that unit roots might be present in a wide variety of macroeconomic
series in levels or logarithms. This finding made the unit root testing very
popular. Aditionally, the implication that variables should be rendered
stationary by differencing before they are included in an econometric model was
generalised.  Further theoretical developments by Granger and Engle in 1987
\cite{engle87} raised the possibility that two or more integrated, non
stationary time series might be cointegrated, so that some linear combination of
these series could be stationary even though each series were not.

\section{Vector error correction model}\label{sec:vecm}

VECM, developed by \cite{engle87}, is a linear model for I(1) variables that are
cointegrated, see \cite{banerjee1993}. If cointegration exists, variable
differences are stationary and they introduce an error correction term which
adjusts coefficients to bring the variables back to equilibrium.

VECM  is obtained re-writing equation \ref{eq:var} in terms of the new
variable $\Delta\mathbf{y}_t=\mathbf{y}_t-\mathbf{y}_{t-1}$.
The VECM model, expressed in terms those differences, takes the form:
\begin{equation}\label{eq:vec}
\Delta \mathbf{y}_t 
= \boldsymbol{\Omega}\,\mathbf{y}_{t-1}
  + \sum_{i=1}^{p-1} \boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
where the coefficients matrices $\boldsymbol{\Phi}_i^*$ and 
$\boldsymbol{\Omega}$, expressed in terms of the matrices
$\boldsymbol{\Phi}_i$ of the model VAR shown in equation \ref{eq:var}, are:
\begin{align*}
\boldsymbol{\Phi}_i^* 
&:= -\sum_{j=i+1}^{p}\boldsymbol{\Phi}_j\,, \\
\boldsymbol{\Omega}
&:= -\left( \mathbb{I} - \boldsymbol{\Phi}_1 - \dots 
    - \boldsymbol{\phi}_p \right)\,. 
\end{align*}
The following well known properties of the matrix $\boldsymbol{\Omega}$
\cite{johansen1995} will be useful in the sequel:
\begin{itemize}
\item
If $\boldsymbol{\Omega} = \mathbf{0}$, there is no cointegration.
\item 
If $rank(\boldsymbol{\Omega})=l$, i.e., if $\boldsymbol{\Omega}$ has
full rank, then the time series are not I(1) but stationary.
\item
If $rank(\boldsymbol{\Omega})=r$, $0<r<l$, then there is cointegration
and the matrix $\boldsymbol{\Omega}$ can be expressed as
$\boldsymbol{\Omega}=\boldsymbol{\alpha\beta}^\top$, where $\boldsymbol{\alpha}$
and $\boldsymbol{\beta}$ are
$l\times r$ matrices and
$\text{rank}(\boldsymbol{\alpha})=\text{rank}(\boldsymbol{\beta})=r$.
\item
The columns of $\boldsymbol{\beta}$ contains the cointegration vectors and the rows of
$\boldsymbol{\alpha}$ correspond with the adjusted vectors. 
$\boldsymbol{\beta}$ is obtained by Johansen procedure~\cite{johansen1988},
whereas $\boldsymbol{\alpha}$ has to be determined as a variable in the VECM.
\end{itemize}
It is worth noticing that the factorisation of the matrix
$\boldsymbol\Omega$ is not unique, since for any $r \times r$
nonsingular matrix $\mathbf{H}$, $\boldsymbol{\alpha}^*:=\boldsymbol{\alpha}\mathbf{H}$,
and $\boldsymbol{\beta}^*=\boldsymbol{\beta}(\mathbf{H}^{-1})^\top$ we have
$\boldsymbol{\alpha\beta}^\top=\boldsymbol{\alpha}^*(\boldsymbol{\beta}^*)^\top$.
If cointegration exists, then equation (\ref{eq:vec}) can be written
as follows:
\begin{equation}\label{eq:vecfull}
\Delta\mathbf{y}_t 
= \boldsymbol{\alpha\beta}^\top\mathbf{y}_{t-1} 
  + \sum_{i=1}^{p-1}\boldsymbol{\Phi}_i^*\,\Delta\mathbf{y}_{t-i}
  + \mathbf{c} + \boldsymbol{\epsilon}_t\,,
\end{equation}
\noindent
which is a VAR model but for time series differences.

@@

Taking into account the error term $\mathbf{E}$, equation~\ref{eq:simpleveceq}
can be solved with respect to $\mathbf{X}$ using the ordinary least
squares estimation (see section \ref{sec:OLS}).
\section{Cointegration Tests}

\section{Ordinary Least Squares method} \label{sec:OLS}

@@

\newpage
